---
title: 'Lesson 5: Introduction to the vector handling in R'
author: "Jan Verbesselt and Sytze de Bruyn"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 4
    number_sections: true
  pdf_document:
    toc: yes
bibliography: refs.bib  
---

# Today's items

## ToDo in the morning

Go through the whole lesson and try to answer the questions below. We will address the questions in the lesson and discuss.

## Objective of today

Learn how to handle vector data (e.g. ESRI shape files)

## Learning outcomes of today:

In today's lecture, we will explore the basics of handling spatial vector data in R. There are several R packages for this purpose but we will focus on using `sp`, `rgdal`, `rgeos` and some related packages. At the end of the lecture, you should be able to:

  - create point, line and polygon objects from scratch;
  - explore the structure of `sp` classes for spatial vector data;
  - plot spatial vector data;
  - transform between datums and map projections; 
  - apply basic operations on vector data, such as buffering, intersection and area calculation;
  - write spatial vector data to a kml file;

# Vector R basics

## Some packages for working with spatial vector data in R

The packages `sp` and `rgdal` are widely used throughout this course. 
Both packages not only provide functionallity for raster data but also for vector data.
The `sp` package provides classes for importing, manipulating and exporting spatial data in R, and methods 


For example, `rgdal` includes bindings to parts of the **OGR Simple Feature Library** which provides access to a variety of vector file formats such as ESRI Shapefiles and kml. The OGR library is part of the widely used **Geospatial Data Abstraction Library** (GDAL). The GDAL library is the most useful freely-available library for reading and writing geospatial data.
The GDAL library is well-documented (http://gdal.org/), but with a catch for R and Python programmers. The GDAL (and associated OGR) library and command line tools are all written in C and C++. Bindings are available that allow acces from a variety of other languages including `R`, `Python` but the documentation is all written for the C++ version of the libraries. This can make reading the documentation rather challenging. Fortunaly, the `rgdal` package, providing GDAL bindings in R, is well documented with lot's of examples. The same is valid for the Python libaries.

Similarly, `rgeos` is an interface to the powerful **Geometry Engine Open Source (GEOS)** library for all kind of operations on geometries (buffering, overlaying, area calculations, etc.).  

Thus, functionality that you commonly find in expensive GIS software is also available within R, using free but very powerful software libaries. 

The possiblities are huge; in this course we can only scratch the surface with some essentials which hopefully invite you to experiment further and use them in your research. Details can be found in the book Applied Spatial Data Analysis with R and several vignettes authored by Roger Bivand, Edzer Pebesma and Virgilio Gomez-Rubio. Owing to time constraints, this lecture cannot cover the related package `spacetime` with classes and methods for spatio-temporal data [@Bivand:2013ux].

## Creating and manipulating geometries

The package `sp` provides classes for spatial-only geometries, such as `SpatialPoints` (for points), and combinations of geometries and attribute data, such as a `SpatialPointsDataFrame`. The following data classes are available for spatial vector data [@Edzer:2005ux]:

| data type     | class                     | attribute     |
|-------------  |-----------------          |-------------  |
| points        | *SpatialPoints*           | No            |
| points        | *SpatialPointsDataFrame*  | data.frame   |
| line          | *Line*                    | No           |
| lines         | *Lines*                   | No           |
| lines         | *SpatailLines*            | No           |
| lines         | *SpatialLinesDataFrame*   | data.frame      |
| rings         | *Polygon*                    | No           |
| rings         | *Polygons*                   | No           |
| rings         | *SpatailPolygons*            | No           |
| rings         | *SpatialPolygonsDataFrame*   | data.frame   |

Table:  Overview of *sp* package classes spatial-only geometries.

We will go through a few examples of creating geometries from scratch to familiarize yourself with these classes.

First, start Google Earth on your computer and make a note of the longitude and latitude of two points in Wageningen that are relevant to you. Use a decimal degree notation with at least 4 digits after the decimal point. To change the settings in Google Earth click *Tools | Options* and change the Show lat/Long setting on the 3D View Tab. 

### Points: SpatialPoints, SpatialPointsDataFrame
The example below shows how you can create spatial point objects from these coordinates. Type `?function name` (e.g. `?cbind`) for finding help on the functions used.

*Question* See ?CRS and have a look at the help.
```{r, message=FALSE}
# load sp package
library(sp)
library(rgdal)

# coordinates of two points identiefied in Google Earth, for example
pnt1_xy <- cbind(5.6660, 51.9872)   # enter your own coordinates
pnt2_xy <- cbind(5.6643, 51.9668)   # enter your own coordinates

# combine coordinates in single matrix
coords <- rbind(pnt1_xy, pnt2_xy)

# make spatial points object
prj_string_WGS <- CRS("+proj=longlat +datum=WGS84")
mypoints <- SpatialPoints(coords, proj4string=prj_string_WGS)
```

```{r eval=FALSE}
# inspect object
class(mypoints)
str(mypoints)
```

```{r, eval=TRUE}
# create and display some attribute data and store in a data frame
mydata <- data.frame(cbind(id = c(1,2), 
                Name = c("my first point", 
                         "my second point")))

# make spatial points data frame
mypointsdf <- SpatialPointsDataFrame(
  coords, data = mydata, 
  proj4string=prj_string_WGS)
```

```{r, eval=FALSE}
class(mypointsdf) # inspect and plot object
names(mypointsdf)
str(mypointsdf)
```

```{r, eval=TRUE}
spplot(mypointsdf, zcol="Name", col.regions = c("red", "blue"), 
       xlim = bbox(mypointsdf)[1, ]+c(-0.01,0.01), 
       ylim = bbox(mypointsdf)[2, ]+c(-0.01,0.01),
       scales= list(draw = TRUE))
```

```{r, eval=FALSE}
## play with the spplot function
## What is needed to make the following work?
spplot(mypointsdf, col.regions = c(1,2))
```

**Question:** What is the the difference between the objects mypoints and mypointsdf.

### Lines

Now let us connect the two points by a straight line. First find information on 
the classes for lines that are available in `sp`. The goal is to create 
SpatialLinesDataFrame but we have to go through some other classes.

```{r}
# consult help on SpatialLines class
(simple_line <- Line(coords))
(lines_obj <- Lines(list(simple_line), "1"))
(spatlines <- SpatialLines(list(lines_obj), proj4string=prj_string_WGS))
(line_data <- data.frame(Name = "straight line", row.names="1"))
(mylinesdf <- SpatialLinesDataFrame(spatlines, line_data))
```

*Q: What is the difference between Line and Lines?*
```{r, echo=FALSE, eval=FALSE}
class(simple_line)
class(lines_obj)
```

```{r, eval=FALSE}
class(mylinesdf)
str(mylinesdf)
```

```{r, fig.height=6, fig.width=12}
spplot(mylinesdf, col.regions = "blue", 
       xlim = bbox(mypointsdf)[1, ]+c(-0.01,0.01), 
       ylim = bbox(mypointsdf)[2, ]+c(-0.01,0.01),
       scales= list(draw = TRUE))
```

*Try to understand the above code and its results by studying help.*
*Try to add the points together with the lines on the same map*


### Writing and reading spatial vector data using OGR

What now follows is a brief intermezzo before we continue with the classess for polygons. 
Let us first export the objects created as kml files that can be displayed in Google Earth. 
We will use the `OGR` functionality available through the package `rgdal`.

```{r, eval=FALSE, echo=TRUE}
library(rgdal)
# write to kml ; below we assume a subdirectory data within the current 
# working directory.
if (!file.exists("data")) { 
  dir.create("data") 
  writeOGR(mypointsdf, file.path("data","mypointsGE.kml"), 
           "mypointsGE", driver="KML", overwrite_layer=TRUE)
  writeOGR(mylinesdf, file.path("data","mylinesGE.kml"), 
           "mylinesGE", driver="KML", overwrite_layer=TRUE)
 }
```

Check (in Google Earth) whether the attribute data were written to the kml output.

The function `readOGR` allows reading OGR compatible data into a suitable Spatial vector object. 
Similar to `writeOGR`, the function requires entries for the arguments `dsn` (data source name) and `layer` (layer name). The interpretation of these enties vary by driver. Please study details in the help file.

Digitize a path (e.g. a bicycle route) between the two points of interest you selected earlier in Google Earth. This can be achieved using the `Add Path` functionality of Google Earth (see [here](https://support.google.com/earth/answer/148072?hl=en) for more info. Save the path in the data folder within the working directory under the name `route.kml`. We will read this file into a spatial lines object and add it to the already existing `SpatialLinesDataFrame` object.

```{r, eval=FALSE}
dsn = file.path("data","route.kml")
ogrListLayers(dsn) ## to find out what the layers are
myroute <- readOGR(dsn, layer = ogrListLayers(dsn))
```

```{r, eval=TRUE, echo=FALSE}
if (!file.exists(file.path("data","route.rda"))) {
  myroute <- readOGR(file.path("data","route.kml"), ogrListLayers(dsn))
  save(myroute, file = file.path("data","route.rda"), compress="xz")
} else {
  load(file.path("data","route.rda"))
}
```

```{r}
# put both in single data frame
myroute@proj4string <- prj_string_WGS
names(myroute)
myroute$Description <- NULL # delete Description
mylinesdf <- rbind(mylinesdf, myroute)
```

Try to understand the above code and results. Feel free to display the data and export to Google Earth.

### Transformation of coordinate system

Transformations between coordinate systems are crucial to many GIS applications. 
The **Keyhole Markup Language (kml)** used by Google Earth uses latitude and longitude in a polar WGS84 coordinate system (i.e. geographic coordinates). However, in some of the examples below we will use metric distances (i.e. carthographic coordinates).There are two types of coordinate systems that you need to recognise: **projected coordinate systems** and **unprojected coordinates systems**

One of the challenges of working with geo-spatial data is that geodetic locations (points on the Earth surface) are mapped into a two-dimensional cartesian plane using a cartographic projection. Projected coordinates are coordinates that refer to a point on a two-dimensional map that *represents* the surface of the Earth (i.e. **projected coordinate system**).Latitude and Longitude values are an example of an **unprojected coordinate system**. These are coordinates that directly refer to a point on the Earth's surface. One way to deal with this is by transforming the data to a planar coordinate system. In `R` this can be achieved via bindings to the **PROJ.4 - Cartographic Projections Library** (http://trac.osgeo.org/proj/), which are available in `rgdal`.  

Central to spatial data in the `sp` package is that they have a coordinate reference system, which is coded in object of `CRS` class. Central to operations on different spatial data sets is that their coordinate reference system is compatible (i.e., identical). This CRS can be a character string describing a reference system in a way understood by the `PROJ.4` projection library, or a (character) missing value. An interface to the `PROJ.4` library is available only if the R package `rgdal` is present.

We will transform our spatial data to the Dutch grid (Rijksdriehoekstelsel), often referred to as RD.
Please note that: 

- some widely spread definitions of the Dutch grid (**EPSG: 28992**) are incomplete 
(see e.g. http://www.spatialreference.org are search for the EPSG number);
- the transformation used below is approximate. 
Details can be found at http://nl.wikipedia.org/wiki/Rijksdriehoekscoordinaten.
- The PROJ.4 details can be found here: http://www.spatialreference.org/ref/epsg/28992/proj4/

```{r}
# define CRS object for RD projection
prj_string_RD <- CRS("+proj=sterea +lat_0=52.15616055555555 +lon_0=5.38763888888889 
+k=0.9999079 +x_0=155000 +y_0=463000 +ellps=bessel +towgs84=565.2369,50.0087,465.658,
-0.406857330322398,0.350732676542563,-1.8703473836068,4.0812 +units=m +no_defs")

# perform the coordinate transformation from WGS84 to RD
mylinesRD <- spTransform(mylinesdf, prj_string_RD)
```

You can always plot the line using the basic plot command

```{r plotmylinesRD, fig.width=3, fig.height=3, message=TRUE, eval=TRUE, fig.align='center',strip.white=TRUE}
plot(mylinesRD, col = c("red", "blue"))
box()
```

Now that the geometries are projected to a planar coordinate system 
the length can be computed using a function from the package rgeos.

```{r, eval=TRUE}
# use rgeos for computing the length of lines 
library(rgeos)
(mylinesdf$length <- gLength(mylinesRD, byid=T))
```

Feel free to export the updated lines to Google Earth or to inspect the contents of the `data` slot of the object `mylinesdf` 

```{r, eval=FALSE}
mylinesdf@data
```


### Polygons

#### Polygons with the `sp` package
We now continue with `sp` classes for polygon objects. The idea is to illustrate the classes; the data are meaningless. Let us create overlapping circles around the points you defined earlier.

```{r circlepoly, eval=TRUE}
# perform the coordinate transformation from WGS84 (i.e. not a projection) to RD (projected)"
# this step is necessary to be able to measure objectives in 2D (e.g. meters)
(mypointsRD <- spTransform(mypointsdf, prj_string_RD))
pnt1_rd <- coordinates(mypointsRD)[1,]
pnt2_rd <- coordinates(mypointsRD)[2,]

# make circles around points, with radius equal to distance between points
# define a series of angles going from 0 to 2pi
ang <- pi*0:200/100
circle1x <- pnt1_rd[1] + cos(ang) * mylinesdf$length[1]
circle1y <- pnt1_rd[2] + sin(ang) * mylinesdf$length[1]
circle2x <- pnt2_rd[1] + cos(ang) * mylinesdf$length[1]
circle2y <- pnt2_rd[2] + sin(ang) * mylinesdf$length[1] 
c1 <- cbind(circle1x, circle1y)
c2 <- cbind(circle2x, circle2y)
```

You can plot everything again using basic R plot commands:

```{r plotcircle, fig.width=4, fig.height=6, message=TRUE, eval=TRUE, fig.align='center'}
plot(c2, pch = 19, cex = 0.2, col = "red", ylim = range(circle1y, circle2y))
points(c1, pch = 19, cex = 0.2, col = "blue")
points(mypointsRD, pch = 3, col= "darkgreen")
```

Now, we create subsequently

- polygons
- SpatialPolygons
- SpatialPolygonsDataFrame

```{r}
# Iterate through some steps to create SpatialPolygonsDataFrame object
circle1 <- Polygons(list(Polygon(cbind(circle1x, circle1y))),"1")
circle2 <- Polygons(list(Polygon(cbind(circle2x, circle2y))),"2")
spcircles <- SpatialPolygons(list(circle1, circle2), proj4string=prj_string_RD)
circledat <- data.frame(mypointsRD@data, row.names=c("1", "2"))
circlesdf <- SpatialPolygonsDataFrame(spcircles, circledat)
```

Similar results can be obtained using the function gBuffer of the package rgeos, as demonstrated below. Notice the use of two overlay functions from the package `rgeos`.

The final results can be plotted using basic R plotting commands:

```{r, fig.height=6, fig.width=4, fig.cap="plot example of final results", fig.align='center'}
plot(circlesdf, col = c("gray60", "gray40"))
plot(mypointsRD, add = TRUE, col="red", pch=19, cex=1.5)
plot(mylinesRD, add = TRUE, col = c("green", "yellow"), lwd=1.5)
box()
```

Here is an example of a plot of the results which employs a few more advanced options of `spplot`.

```{r spplotresult, fig.height=6, fig.width=4, fig.cap="spplot example", fig.align='center'}
spplot(circlesdf, zcol="Name", col.regions=c("gray60", "gray40"), 
       sp.layout=list(list("sp.points", mypointsRD, col="red", pch=19, cex=1.5), 
                      list("sp.lines", mylinesRD, lwd=1.5)))
```

Try to understand how `spplot` works by breaking it down in simple steps e.g.

```{r, eval=FALSE}
spplot(circlesdf, zcol="Name", col.regions=c("gray60", "gray40"))
```

**Question: **  Which plotting options do you prefer? E.g `spplot` or `plot`

#### Polygon operations with `rgeos` (buffer, intersect, difference)

```{r buffering, eval = TRUE, fig.width=4, fig.height=4, fig.align='center'}
library(rgeos)
## Expand the given geometry to include the area within the specified width with specific styling options.
buffpoint <- gBuffer(mypointsRD[1,], width=mylinesdf$length[1], quadsegs=2)
mydiff <- gDifference(circlesdf[1,], buffpoint)

plot(circlesdf[1,], col = "red")
plot(buffpoint, add = TRUE, lty = 3, lwd = 2, col = "blue")

gArea(mydiff) ## what is the area of the difference?

plot(mydiff, col = "red")

myintersection <- gIntersection(circlesdf[1,], buffpoint)

plot(myintersection, col="blue")

gArea(myintersection)
print(paste("The difference in area =", round(100 * gArea(mydiff) / 
                                             gArea(myintersection),2), "%"))

```

- **Question ** What happens if you change `quadsegs` to a higher number?
- **Question ** Do you understand the script? What is the difference between `gIntersection` and `gDifference`?

# Today's summary

We learned about:

- projection systems: unprojected versus projected
- GDAL/OGR and GEOS C/C++ libraries with binding in R via packages `rgdal`, `rgeos`.
- 

# Excercise of today

Download the following data set from:

https://www.dropbox.com/s/w25dwfo7c8wuaxv/read-write-shapefiles.zip?dl=0

- Read in the following files using the `rgdal` library 
- create a map that visualise the rivers, counties, and centroids using R
- write the files to kml files

```{r, eval=FALSE}
ogrInfo(dsn, "nw-rivers")
ogrInfo(dsn, "nw-centroids")
ogrInfo(dsn, "nw-counties")
```

Questions: 

- Where are these spatial vector objects geographically located?

## Optional for bonus points

- What needs to be done to derive a buffer around the rivers of 200m?

# More info

About projections and code: https://www.nceas.ucsb.edu/scicomp/recipes/projections

# References